\vspace{-1cm}\normalsize{\textbf{Fairly Random: Impact of Winning the Toss on the Probability of Winning}}

### 2016-03-06 write_up/cricket.tex
cricket draft

\begin{document}
\title{\vspace{-.5cm}\normalsize{Fairly Random: The Impact of Winning the Toss on the Probability of Winning\footnote{You can find the scripts used to scrape the data, the final data set, and scripts used for analysis at: \href{https://github.com/dwillis/cricket-stats}{https://github.com/dwillis/cricket-stats}.}\vspace{.5cm}}}
\author{\normalsize{Gaurav Sood}\\\href{mailto:gsood07@gmail.com}{\small{gsood07@gmail.com}} \and \normalsize{Derek Willis}\\\href{mailto:dwillis@gmail.com}{\small{dwillis@gmail.com}}}
\date{\vspace{.5cm}\normalsize{\today}}
\maketitle
\doublespacing

@@

But, rather than consider all home matches, we may instead want to only consider matches that are officiated only by home umpires --- the norm till 1992.\footnote{For more information on move to neutral umpires, see \href{http://www.espncricinfo.com/magazine/content/story/511175.html}{Neutral Umpires} by S. Rajesh on ESPNcricinfo.} In matches featuring umpires from only the home country, the team with `home umpire' advantage won toss nearly 51.9\% of the times. And the chance of getting a greater percentage of wins than 51.9\% in 2,965 is a shade less than 4\%. Thus, there is some reason to worry that the tosses are rigged. Any such rigging would bias estimates of advantage of winning the coin toss to the extent that it is correlated with ability. More plainly, if stronger teams win more tosses, estimates of the advantage of the coin toss would be inflated upwards. And vice versa, if otherwise. We, however, do not have good reasons to think that there is a correlation. So for now, we proceed as if the tosses are random.

@@

The conventional wisdom among lay cricket followers is that toss grants the greatest advantage in multi-day affairs like tests and first class matches, followed by day long affairs, and Twenty20s. And there is good dose of common sense behind the conventional wisdom. Pitches invariably deteriorate over multiple days and batting last in a test match is often the most challenging time to bat. The pitch deteriorates far less over the course of the day, or in case of Twenty20s, a few hours. And indeed unlimited over matches provide the most consistent advantage --- the average advantage over FC and test matches is north of 2.7\% (see Figure~\ref{fig:type}). Looked in relative terms, the advantage of winning the toss is also close to the greatest in multi-day affairs. Only about 60\% of test matches end in a  clear decision, the rest end in a draw. Thus, the advantage is closer to 4.5\%.

The heftiest advantage, however, is in domestic one-day matches (List A), approximately 4.8\%. The number for ODIs, however, is a bit puzzling --- there is an advantage to winning the toss but it is really slight (.83\%). That can best be ascribed to choosing badly. In T20s and T20Is, the advantage is 1.3\% and 3.8\% respectively.

\begin{figure}[htbp]
\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Type of Match.}
\begin{minipage}{0.85\textwidth}
\centering
\includegraphics[scale=.85]{../figs/winbyType.pdf}
{\footnotesize Note: \emph{n} refers to the number of matches.\par}
\end{minipage}
\label{fig:type}
\end{figure}

But type of matches cover only major source of variation and theorizing about the advantage granted by the toss. It is often claimed that the toss is more crucial in day and night matches. Due to dew --- it is thought to make bowling hard, and the visibility of the white ball is thought to be lower under lights, which makes catching hard --- it is thought that the team that fields second is disadvantaged. It turns out that the conventional wisdom is largely vindicated, except perhaps for Twenty20 Internationals (but we have very little on Day and Night T20 Internationals to say anything with great certainty) (see Figure ~\ref{fig:dn}). In each, domestic one-day and Twenty20 matches, and one-day internationals, the advantage of winning the toss in Day and Night matches is heftier, and in the case of ODIs and domestic T20s, markedly so --- the difference in advantage of winning the toss between Day and Night and Day versions is over 9\% for both. Splitting by whether or not the match spans Day and Night also reveals that in day ODIs, the team that wins the toss loses more often than the team that wins it. We can  attribute that to choosing badly.

\begin{figure}[htbp]
\centering
\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Day or Day and Night}
\includegraphics[scale=.85]{../figs/winbyDayNight.pdf}
\label{fig:dn}
\end{figure}

In limited over matches sometimes weather intervenes significantly after the match has already begun. After certain loss of time, the match is generally curtailed and the scores adjusted using a method invented by Duckworth and Lewis \citep[see][]{duckworth1998}. We can use the random nature of who wins the toss to see if winning percentages are strongly conditioned by matches that use Duckworth-Lewis and matches that don't. Again the conventional wisdom suggests that the team that wins the toss is at an advantage in such matches. And again, the conventional wisdom is vindicated. In List A, ODI and domestic Twenty20 matches, the advantage of winning the toss is greater when Duckworth Lewis is used to adjust the scores than when it isn't --- the difference in advantage ranges from about .75\% to over 2\% (see Figure ~\ref{fig:dl}).

\begin{figure}[htbp]
\centering
\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Whether or not Duckworth-Lewis was invoked}
\includegraphics[scale=.85]{../figs/winbyDL.pdf}
\label{fig:dl}
\end{figure}

Winning the toss ought to matter the most when the difference between the quality of the teams that are playing is the least. Similarly, it is unlikely that winning the toss would change who wins the game when two ill-matched teams are playing. To study this, we collected data on team quality. The ICC publishes team ratings for international test and one-day teams each month.\footnote{For details about how the ICC produces these ratings, see \href{http://icc-live.s3.amazonaws.com/cms/media/about_docs/536b1a48c16e5-Reliance\%20ICC\%20ODI\%20Team\%20Rankings\%20FAQs\%202014.pdf}{ICC Rating FAQs}.} Ratings of the ODI teams have been published since 1981, and of test teams, since 1952. Of the entire dataset that spans 1981-today and 1952-today, we only have we have data till 2013.\footnote{The format in which the ICC publishes the ratings changed in 2013. And scraping the latter data posed additional hurdles. We decided that the additional effort wasn't worth the small amount of additional data.}

Team ratings range from 0 to 143 in our data. For instance, Bangladesh had a rating of 0 in tests for most of 2002 and 2003. And Australia in 2007 twice held a rating of 143. Both seem reasonable. We measure how closely matched the teams by differencing the ranking points of one team from the other. Commercial considerations mean that a majority of the games are played among highly ranked and closely matched teams. So, we limit our analysis to matches where teams ratings differ by no more than 30 points.

The results are expected, but new. As Figure ~\ref{fig:ranks}, there is a sharp curve around 0. When closely matched teams win, the team that wins the toss has a consistent advantage over the team that loses the toss.

\begin{figure}[htbp]
\centering
\caption{Percentage of Matches Won Minus Matches Lost After Winning the Toss by Difference in Ranks}
\includegraphics[width=1\textwidth]{../figs/winbyRank.pdf}
\label{fig:ranks}
\end{figure}

How the advantage vary by country? Are some countries better than others? We investigated the question by tallying the advantage by team. Sri Lanka and India do especially well (see ~\ref{fig:country}), coming in at 4.53\% and 3.35\% respectively. Australia and West Indies hover around 1.3--1.5\% and England at 2.33\%. New Zealand, however, does worse when it wins the toss than when it loses it --- difference is about 1.21\%.

\begin{figure}[htbp]
\centering
\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Country}
\includegraphics[scale=.85]{../figs/winbyCountry.pdf}
\label{fig:country}
\end{figure}

This covers some of the more straightforward questions. Other more nuanced but commonly thought questions remain. For one, students of the game suspect that the advantage of winning a toss in early English season is especially great. We next check that using data on International matches from England.

\begin{figure}[htbp]
\centering
\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Month in England}
\includegraphics[scale=.85]{../figs/winbyMonthEngland.pdf}
\label{fig:season}
\end{figure}

This isn't a comprehensive inventory of analyses that one could do on the topic. But neither is it meant to be. Some simple questions still remain unanswered. For one, how does the toss advantage vary over time? Have teams become smarter?  We propose to investigate these questions in a follow-up piece.

>>>>

\begin{document}
\title{\vspace{-.5cm}\normalsize{Fairly Random: The Impact of Winning the Toss on the Probability of Winning\footnote{The paper benefitted from comments by Andrew Gelman, Donald Green, and Daniel Stone. Scripts used to scrape the data, the final data set, and scripts used for analysis can be found at: \href{https://github.com/dwillis/cricket-stats}{https://github.com/dwillis/cricket-stats}. }\vspace{.5cm}}}
\author{\normalsize{Gaurav Sood}\\\href{mailto:gsood07@gmail.com}{\small{gsood07@gmail.com}} \and \normalsize{Derek Willis}\\\href{mailto:dwillis@gmail.com}{\small{dwillis@gmail.com}}}
\date{\vspace{.5cm}\normalsize{\today}}
\maketitle
\doublespacing

@@

However, rather than consider all home matches, we may instead want to only consider matches that are officiated only by home umpires---the norm till 1992.\footnote{For more information on move to neutral umpires, see \href{http://www.espncricinfo.com/magazine/content/story/511175.html}{Neutral Umpires} by S. Rajesh on ESPNcricinfo.} In matches featuring umpires from only the home country, the team with `home umpire' advantage won toss nearly 51.9\% of the times. And the chance of getting a greater percentage of wins than 51.9\% in 2,965 is a shade less than 4\%. Thus, there is some reason to worry that the tosses are rigged. Any such rigging would bias estimates of advantage of winning the coin toss to the extent that it is correlated with ability. More plainly, if stronger teams win more tosses, estimates of the advantage of the coin toss would be inflated upwards. And vice versa, if otherwise. We, however, do not have good reasons to think that there is a correlation. So for now, we proceed as if the tosses are random.

@@

The conventional wisdom among lay cricket followers is that toss grants the greatest advantage in multi-day affairs like tests and first class matches, followed by day long affairs, and Twenty20s. And there is good dose of common sense behind the conventional wisdom. Pitches invariably deteriorate over multiple days and batting last in a test match is often the most challenging time to bat. The pitch deteriorates far less over the course of the day, or in case of Twenty20s, a few hours. And indeed unlimited over matches provide the greatest advantage--- the average advantage over FC and test matches is north of 2.6\% (see Figure~\ref{fig:type}). Looked in relative terms, the advantage of winning the toss is also close to the greatest in multi-day affairs. Only about 60\% of test matches end in a clear decision, the rest end in a draw. Thus, the advantage is closer to 4.5\%. The heftiest raw advantage, however, is in one-day matches (List A and ODIs), approximately 3.3\%. In T20s and T20Is, the advantage is considerably smaller, just about 1.27\%. And unlike the estimate of advantage for multi-day and one-day affairs, we cannot statistically reject that the possibility that there is no advantage.

\begin{figure}[htbp]
\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Type of Match.}
\begin{minipage}{0.85\textwidth}
\centering
\includegraphics[scale=.85]{../figs/winbyType.pdf}
{\footnotesize Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
\end{minipage}
\label{fig:type}
\end{figure}

Type of matches are but one source of variation and theorizing about the advantage granted by the toss. It is often claimed that the toss is more crucial in day and night matches. Due to dew---it is thought to make bowling hard, and the visibility of the white ball is thought to be lower under lights, which makes catching hard---the team that fields second is thought to be at a disadvantage. The conventional wisdom is largely vindicated for one day affairs (see Figure ~\ref{fig:dn}). In one-day matches, the advantage of winning the toss in a day and night match is 5.92\%, whereas the advantage of winning the toss in a one-dayer played during the day is less than half---2.89\%. In Twenty20s---domestic and international---, however, we cannot distinguish between the advantage granted by the toss in day and night matches and day time affairs.

\begin{figure}[htbp]
\centering
\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Day or Day and Night}
\includegraphics[scale=.85]{../figs/winbyDayNight.pdf}
{\footnotesize Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
\label{fig:dn}
\end{figure}

Weather has a large impact on the playing conditions in cricket. For instance, cool overcast weather is thought to aid swing bowling, especially on certain pitches. More generally, the advantage of winning the toss likely varies by weather. However, we do not have data on weather. But, we can proxy it with seasons. In particular, students of the game suspect that the advantage of winning the toss in early English season is especially great. We next assess whether that is so.

There is some evidence of a seasonal pattern, with advantage of winning the toss somewhat greater in spring and early summer (May and June) than in mid and late summer and early fall (July to September) ~\ref{fig:season}. However, the thing that catches attention is the large disadvantage of winning the toss in April. We don't have a good explanation for the pattern, except for teams choosing badly.

\begin{figure}[htbp]
\centering
\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Month in England}
\includegraphics[scale=.85]{../figs/winbyMonthEngland.pdf}
{\footnotesize Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
\label{fig:season}
\end{figure}

Aside from affecting the playing conditions, weather affects cricket matches in other, more forceful ways---interrupting, and sometimes ending matches. When a limited over match that is already underway is interrupted by bad weather, and more than a certain amount of time is lost due to the interruption, the match is curtailed and the total that the team batting second must achieve to win is adjusted using a method invented by Duckworth and Lewis \citep[see][]{duckworth1998}.\footnote{Before the Duckworth and Lewis method was adopted, weather affected matches sometimes continued on the next day; an extra day was deliberately left in the schedule for dealing with such eventualities. In cases where the spare day proved inadequate, the match was declared a draw.}

We can use the random nature of who wins the toss to see if winning percentages of the teams that win the toss are strongly conditioned by whether or not Duckworth-Lewis is used. If the advantage of winning the toss in matches using Duckworth-Lewis is different from matches that don't use it, it suggests that the Duckworth-Lewis method is biased. (For a precise estimate of the bias, ideally, we would want to compare matches using Duckworth-Lewis method with matches held in similar conditions.)

In both one-day and Twenty20 matches, the advantage of winning the toss in a match where the target is adjusted using Duckworth-Lewis, is considerably greater (see Figure ~\ref{fig:dl}). In one-day matches, the advantage is 5.35\% in matched adjudicated by Duckworth-Lewis and 3.17\% in matches that don't use it. Statistically, the chances that the two numbers are the same is less than 10\%. (And if you make the plausible assumption that winning a toss only improves the chances of winning, the chance that the two numbers are the same is half that.) In Twenty 20s, the advantage of winning the toss shrinks from 3.9\% in matches using Duckworth-Lewis to 1.17\% in matches without it. Once again, the chance that the two numbers are the same is about 10\%.

\begin{figure}[htbp]
\centering
\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Whether or not Duckworth-Lewis was invoked}
\includegraphics[scale=.85]{../figs/winbyDL.pdf}
{\footnotesize Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
\label{fig:dl}
\end{figure}

Winning the toss ought to matter the most when the difference between the quality of the teams that are playing is the least. Similarly, it is unlikely that winning the toss would change who wins the game when two ill-matched teams are playing. To study this, we collected data on team quality. The ICC publishes team ratings for international test and one-day teams each month.\footnote{For details about how the ICC produces these ratings, see \href{http://icc-live.s3.amazonaws.com/cms/media/about_docs/536b1a48c16e5-Reliance\%20ICC\%20ODI\%20Team\%20Rankings\%20FAQs\%202014.pdf}{ICC Rating FAQs}.} Ratings of the men's ODI teams have been published since 1981, and of the test teams, since 1952. Of the entire ranking dataset, that spans 1981--today and 1952--today for ODI and test teams respectively, we only have we have data till 2013.\footnote{The format in which the ICC publishes the ratings changed in 2013. And scraping the latter data posed additional hurdles. We decided that the additional effort wasn't worth the small amount of additional data.}

Team ratings range from 0 to 143 in our data. For instance, Bangladesh had a rating of 0 in tests for most of 2002 and 2003. And Australia in 2007 twice held a rating of 143. We measure how closely matched the teams by differencing the ranking points of one team from the other. Commercial considerations mean that a majority of the games are played among highly ranked and closely matched teams. Thus, the precision of our estimates is greatest for matches between closely matched teams.

The results are expected, but new. As Figure ~\ref{fig:ranks} --- which plots percentage of matches won or drawn by the team that won the toss--- illustrates, there is a sharp curve around 0. When closely matched teams win, winning the toss has a large impact on the probability of winning.

\begin{figure}[htbp]
\centering
\caption{Percentage of Matches Won Minus Matches Lost After Winning the Toss by Difference in Ranks}
\includegraphics[width=1\textwidth]{../figs/winbyRank.pdf}
{\footnotesize Note: Smoothed relationship between difference in ranks and winning probability by whether or not the team won the toss.\par}
\label{fig:ranks}
\end{figure}

Lastly, we investigate how advantage varies by country in international matches? Are some countries better than others? We investigated the question by tallying the advantage by team. All of our estimates are imprecise enough that we cannot discount the possibility that the true effect is zero. The most puzzling result is from New Zealand, which seems to do more poorly in matches where it wins tosses than where it loses it. On the other hand, Sri Lanka and India do especially well (see ~\ref{fig:country}), coming in at 3.85\% and 3.09\% respectively. Australia, Pakistan, and West Indies hover around 1--1.5\% and England is at 2.29\%.

\begin{figure}[htbp]
\centering
\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Country}
\includegraphics[scale=.85]{../figs/winbyCountry.pdf}
{\footnotesize Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
\label{fig:country}
\end{figure}

This covers some of the more straightforward questions. However, some simple questions still remain unanswered. For one, how does the toss advantage vary over time? For another, how does winning the toss affect the margin of victory?  We propose to investigate these questions in a follow-up piece.
